Experimentally Informed Site-Specific Substitution Models Substantially Deepen Viral Divergence Estimates

### 2018-01-16 manuscript/manuscript.tex

\newcommand\skhcomment[1]{{\color{magenta}#1}}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Comparison of substitution model features.}
Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
The models YNGKP M0, YNGKP M5, ExpCM, and ExpCM+$\Gamma\omega$ represent all possible combinations of these two features. 
Blue indicates presence and white indicates absence of a feature. 
from \ref{tab:wsn_low_params}
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and rate variation on long branch estimation.}
The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:wsn_low_params}.
The trajectories of the YNKGP M0 (blue) and YNGKP M5 (green) do not vary from panel to panel because neither model is site-specific. 
The deviation in trajectory of the ExpCM (red) from the YNGKP M0 (blue) increases from left to right as the mutational constraint of the amino-acid profile increases (logoplots, above). 
The deviation in trajectory of the YNGKP model with a site-specific $\omega$ value inferred from the ExpCM (yellow, \ref{eq:w_r}) is also positively correlated with the constraint of the site-specific amino-acid profiles but the effect size is smaller. 
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
\caption{\label{simulations}
\textbf{Model performance under simulated, site-specific data.} 
Alignments were simulated under an ExpCM (\ref{tab:sim_params}) along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
The randomized ExpCMs have amino-acid profiles shuffled among the sites 
These randomized models are still site-specific but the relationship between the site and the experimental data is broken. 
Grey points represent the length of one branch and the black points are the mean branch lengths over eight simulations. 
The grey, dashed line is the reference line $y=x$, depicting the behavior of a model which is an unbiased estimator of the simulated branch length. 
}
\end{figure}

@@

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_doud}}
\caption{\label{fig:prefs_doud}
\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.21$)} 
}
\end{figure}

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/prefs_lee}}
\caption{\label{fig:prefs_lee}
\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.46$)} 
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/tree_doud}}
\caption{\label{fig:tree_doud}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
}
\end{figure}

>>>>

\newcommand\skhcomment[1]{{\color{magenta}#1}}
\newcommand\jdbcomment[1]{{\color{red}#1}}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.55\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Comparison of substitution model features.}
\jdbcomment{This figure is too big. Maybe put the three panels side-by-side, increase the space between panels, and increase the font sizes. The logo stacks are kind of tall relative to their width.}
Site-specific amino-acid profiles and $\Gamma$-distributed rate variation are both substitution model features which have been shown or theorized to lengthen branches. 
The models YNGKP M0, YNGKP M5, ExpCM, and ExpCM+$\Gamma\omega$ represent all possible combinations of these two features. 
Blue indicates presence and white indicates absence of a feature. 
from \ref{tab:wsn_low_params}
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.75\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\jdbcomment{I still think text could be bigger. Logo stacks still show trimming, maybe a bit tall? The Model key could probably be to the right or left.}
\textbf{Effect of stationary state and rate variation on long branch estimation.}
The expected pairwise identity trajectories were calculated using \ref{eq:f} and models described in \ref{tab:wsn_low_params}.
The trajectories of the YNKGP M0 (blue) and YNGKP M5 (green) do not vary from panel to panel because neither model is site-specific. 
The deviation in trajectory of the ExpCM (red) from the YNGKP M0 (blue) increases from left to right as the mutational constraint of the amino-acid profile increases (logoplots, above). 
The deviation in trajectory of the YNGKP model with a site-specific $\omega$ value inferred from the ExpCM (yellow, \ref{eq:w_r}) is also positively correlated with the constraint of the site-specific amino-acid profiles but the effect size is smaller. 
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
\caption{\label{simulations}
\textbf{Model performance under simulated, site-specific data.} 
\jdbcomment{Fonts are small, and tick labels definitely too small. I would show averaged.}
Alignments were simulated under an ExpCM (\ref{tab:sim_params}) along an HA tree and the branches were re-optimized by a model from the ExpCM or YNGKP family. 
The randomized ExpCMs have amino-acid profiles shuffled among the sites 
These randomized models are still site-specific but the relationship between the site and the experimental data is broken. 
Grey points represent the length of one branch and the black points are the mean branch lengths over eight simulations. 
The grey, dashed line is the reference line $y=x$, depicting the behavior of a model which is an unbiased estimator of the simulated branch length. 
}
\end{figure}

@@

\begin{figure}[H]
\centerline{\includegraphics[width=0.5\textwidth]{figures/tree_doud}}
\caption{\label{fig:tree_doud}
\textbf{Trees optimized with an ExpCM defined by H1 preferences lengthen branches from the focal H1 sequence compared to YNGKP models.} 
\jdbcomment{Keep branch lengths in substitutions per site. Make a panel A show a prefs snippet, B H1 trees, C H3 trees, D H1+H3. Maybe increase very small font sizes a bit.}
The branch lengths of a base topology inferred using the GTR-CAT model were optimized by \textbf{(A)} an ExpCM defined by H1 preferences, \textbf{(B)} an ExpCM+$\Gamma\omega$ defined by H1 preferences, \textbf{(C)} YNKGP M0, and \textbf{(D)} YNGKP M5.
The branch lengths are normalized to the distance between A/South Carolina/1/1918 and A/Solomon Islands/3/2006 and colored to indicate the distance from the H1 focal sequence (black triangle).
}
\end{figure}

@@

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_doud}}
\caption{\label{suppfig:prefs_doud}
\textbf{H1 preferences measured by \cite{doud2016accurate} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_doud}A  ($\beta = 1.21$)} 
}
\end{suppfig}

\begin{suppfig}[H]
\centerline{\includegraphics[width=\textwidth]{figures/prefs_lee}}
\caption{\label{suppfig:prefs_lee}
\textbf{H3 preferences measured by \textit{lee} rescaled with the ExpCM stringency parameter optimized in \ref{fig:tree_lee}A  ($\beta = 1.46$)} 
}
\end{suppfig}

@@
